P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2025-05-08 sage_latex_template/manuscript.tex
minor
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

The goals of this paper are summarized as follows:
\begin{itemize}
    \item We present our code generator tool P4IRS, describing how it works and how it can be used to generate efficient particle simulation codes.
    \item We discuss the advantages of using P4IRS to achieve performance portability, also comparing it with other previous approaches such as tinyMD (\cite{tinymd}) and MESA-PD (\cite{mesapd1,mesapd2}).
    \item We describe P4IRS interfaces to allow its generated code to be integrated with other simulation tools, specifically in the context of domain-partitioning and communication.
    \item We provide performance results for Intel Ice Lake and AMD Milan CPUs, as well as for NVIDIA A40 and A100 accelerators to demonstrate the state-of-the-art performance of the applications generated by P4IRS in comparison to MD-Bench and MESA-PD.
    \item We display weak-scaling results in the Fritz and JUWELS Booster clusters to demonstrate the weak-scaling capabilities of P4IRS in MD and DEM simulations on modern supercomputers.
\end{itemize}

@@

To evaluate performance on a single CPU core or GPU in this paper, we use MD-Bench and miniMD as the baseline performance, as they are well-known implementations that reach state-of-the-art performance for MD simulations.
For DEM cases, we use MESA-PD as the comparison target, since our focus is to show that we can achieve performance-portability with the flexibility that MESA-PD provides.

@@

The Linked Cells results in computing only forces for particles within a region that is discretized by their current and neighbor cells/bins, which can still results in a significant amount of redundant computations.
In order to avoid even more redundant computations, the Verlet Lists strategy can be used jointly with Linked Cells.
The Verlet Lists (see Figure~\ref{fig:verlet_lists}) constructs a list of neighbor particles for every particle, and this list can be created by using the Linked Cells structure, evaluating the distances among particles within close bins and then using the distance to determine whether such particles should be added into the list.
In the Verlet Lists case, the region for searching the neighbor particles is defined by a sphere, where the center of the sphere is the position of the particle and the radius is the cutoff radius plus a ``skin''.
The ``skin'' is used to avoid building the neighbor-lists at every time-step, it extends the neighbor-search and avoid missing computations from particles that end up entering the cut-off region in subsequent time-steps in which the neighbor-lists are not reconstructed.

@@

Note that this introduces a new kernel to build the neighbor-lists, and this is only beneficial when the force calculation performance improvement compensates the time spent in the new kernel.
Besides, in cases where the code is not properly vectorized (especially in CPU where explicit instructions must be generated), the improvement may not be substantial.
In GPUs, however, this should lead to better execution since threads are not halted waiting for the surrounding neighbor cells to be computed when the amount of particles among them is different.

@@

In order to attain distributed-memory parallelism and software flexibility while attaining good performance, P4IRS provides a customizable and robust communication module that facilitates integrating new domain partitioning schemes into its generated simulations.
This is done by combining its runtime functions and code generation techniques, and allows the usage of the waLBerla domain-partitioning scheme based on a block-forest data structure.

@@

These methods make use of a domain-partitioner class that can be implemented in the P4IRS runtime \RMchange{library} to determine which particles must be sent.
The method is implemented in C++ and fill\RMchange{s} domain buffers that are used for evaluating if a particle is outside the domain or within the border using its position.
This means that to include new domain partitioning schemes in P4IRS, it is only necessary to define a C++ runtime class, which can make use of external libraries and therefore can use the waLBerla block-forest domain-partitioning.
Note that the reason for using buffers instead of evaluating the particle positions directly in the runtime code is performance, since particle packing and unpacking kernels should be executed in the target processor to achieve parallelism and to avoid copying data back and forth from the accelerator and host processors.

@@

Experiments for \ac{MD} use the standard Copper FCC lattice (see \autoref{fig:copper_fcc_lattice}) test-case from miniMD/MD-Bench, and we use our results from MD-Bench as a performance reference since it is a well optimized code for \ac{MD}.
The \ac{DEM} test case is a ``Settling Spheres'' benchmark (see \autoref{fig:settling_spheres}), on which several spheres fall into a ``ground'' (half-space), thus generating a ``bed'' of spherical particles.
Due to its multiple shapes and dynamics that consists of many interactions among spheres and half-spaces, it should provide a good reference for DEM simulations.
We compare our single-core results from P4IRS with MESA-PD using 2 MPI ranks since it is not possible to run this setup in MESA-PD using a single rank.
GPU results are not available for MESA-PD as it is not capable of generating GPU code, but we show GPU performance results for P4IRS comparing different code generation strategies.

We also show weak-scaling results to demonstrate P4IRS scaling capabilities.
For this we use regular domain-partitioning since these experiments do not require sophisticated load-balancing algorithms (Copper lattice is an homogeneous test-case in all dimensions, and Settling Spheres can obtain acceptable weak-scaling by regularly partitioning the domain in X and Y dimensions).
We use double floating-point precision and display results when enabling and disabling Newton's 3rd Law optimization (half neighbor-lists).
For \ac{MD} simulations, we show results with the Verlet Lists optimization strategy since it is widely employed in the field, while for \ac{DEM} we display results with the Linked Cells data-structure because particles represent rigid bodies that need to be in contact for their contributions to be computed.
Since the amount of contacts is typically way less than atomic interactions, there are no significant advantages of using the Verlet Lists data structure.

@@

For \ac{MD} case, we use the Verlet Lists algorithm as it is the standard one for such simulations.
The simulations runs through 200 time-steps, and the neighbor-lists are rebuilt at every 20 time-steps of the simulation (and at the start).
We use the simple \ac{LJ} potential (as shown in \autoref{lst:pairslj}) to calculate the force contributions among atoms.

@@

\autoref{fig:md_weak_scaling_fritz} shows the results for P4IRS weak-scaling in the Fritz supercomputer.
For each node, a system of $128^3$ unit cells is included, with roughly 8.388.608 atoms per node (or 116.508 atoms per CPU core).
Since the \emph{multinode} partition in Fritz has the maximum of 64 nodes (which results in 4608 CPU cores, no hyper-threading), this is the maximum amount we scale up our problem to.
It is possible to see that we achieve perfect weak-scaling in all the settings we use, and the \ac{HN} version is significantly faster than the \ac{FN} version.
Notice that atomic operations are not required because we only make use of parallelism through spatial decomposition and message communication with MPI, thus we use distributed-memory parallelism strategy even within the same node and there is no risk of race conditions.
Based on our results, it is clear that P4IRS is able to generate scalable code on modern CPU clusters, but problems that are not too homogeneous may need proper mechanisms for balancing the workload among different MPI ranks.
This is possible by using different domain-partitioning methods in P4IRS such as the waLBerla domain partitioning with block-forests, but for this example we can achieve perfect scaling even with regular domain-partitioning.

@@

\autoref{fig:md_weak_scaling_juwels} shows the weak-scaling runtimes for P4IRS in the JUWELS Booster cluster.
Each node simulates a system of $100^3$ unit cells, with roughly 500.000 atoms per GPU.
We scale up to 128 nodes, which is equivalent of 512 A100 GPUs.
Note that for the first node, the performance is slightly worse than the next ones, which can be explained as we only have 4 GPUs in one node, and domain-partitioning is only done in two dimensions.
This can lead to worse performance because the surface area for communication is likely higher for the dimension where the partitioning is not done, which culminates in more atoms to be communicated and therefore more time spent packing and transferring data.
Apart from this, the scaling is also perfect for all the settings, and it is possible to observe that the \ac{FN} version is slightly faster than \ac{HN}.
As we already discussed, the usage of \ac{HN} is not necessarily beneficial in GPUs due to atomics, and this can be worse especially for kernels which do not exhibit high arithmetic intensity.
The results demonstrate that P4IRS is in fact capable of generating scalable code for modern GPU clusters.

@@

\autoref{fig:dem_single_core_runtimes} displays the runtimes for the \ac{DEM} kernels in single-core (P4IRS) and two cores/ranks (P4IRS/MESA-PD), also with cases building the neighbor-lists per cell (NPC) in P4IRS to evaluate its impact on CPU performance.
It is noticeable that P4IRS outperforms MESA-PD even with a single rank in the overall runtime, and what draws the attention is the amount of time MESA-PD spends executing other procedures instead of the force calculation and building the neighbor-lists
This can be attributed to the overhead that MESA-PD has to manage its domain-partitioning with waLBerla (which cannot be switched) and calling its communication routines for next-neighbor communication.
A more in-depth analysis is required to actually understand why it takes a significant fraction of the time for MESA-PD to communicate data, but this is out of scope of this paper.

>>>>

The goals of this paper are summarized as follows:
\begin{itemize}
    \item We present our code generator tool P4IRS, describing how it works and how it can be used to generate efficient particle simulation codes.
    \item We discuss the advantages of using P4IRS to achieve performance portability, also comparing it with other previous approaches such as tinyMD (\cite{tinymd}) and MESA-PD (\cite{mesapd1,mesapd2}).
    \item We describe P4IRS interfaces to allow its generated code to be integrated with other simulation tools, specifically in the context of domain partitioning and communication.
    \item We provide performance results for Intel Ice Lake and AMD Milan CPUs, as well as for NVIDIA A40 and A100 accelerators to demonstrate the state-of-the-art performance of the applications generated by P4IRS in comparison to MD-Bench and MESA-PD.
    \item We display weak-scaling results in the Fritz and JUWELS Booster clusters to demonstrate the weak-scaling capabilities of P4IRS in MD and DEM simulations on modern supercomputers.
\end{itemize}

@@

To evaluate performance on a single CPU core or GPU in this paper, we use MD-Bench and miniMD as the baseline performance, as they are well-known implementations that reach state-of-the-art performance for MD simulations.
For DEM cases, we use MESA-PD as the comparison target, since our focus is to show that we can achieve performance portability with the flexibility that MESA-PD provides.

@@

The Linked Cells results in computing only forces for particles within a region that is discretized by their current and neighbor cells/bins, which can still results in a significant amount of redundant computations.
In order to avoid even more redundant computations, the Verlet Lists strategy can be used jointly with Linked Cells.
The Verlet Lists (see Figure~\ref{fig:verlet_lists}) constructs a list of neighbor particles for every particle, and this list can be created by using the Linked Cells structure, evaluating the distances \RMchange{between} particles within close bins and then using the distance to determine whether such particles should be added into the list.
In the Verlet Lists case, the region for searching the neighbor particles is defined by a sphere, where the center of the sphere is the position of the particle and the radius is the cutoff radius plus a ``skin''.
The ``skin'' is used to avoid building the neighbor-lists at every time-step, it extends the neighbor-search and avoid missing computations from particles that end up entering the cut-off region in subsequent time-steps in which the neighbor-lists are not reconstructed.

@@

Note that this introduces a new kernel to build the neighbor-lists, and this is only beneficial when the force calculation performance improvement compensates the time spent in the new kernel.
Besides, in cases where the code is not properly vectorized (especially in CPU where explicit instructions must be generated), the improvement may not be substantial.
In GPUs, however, this should lead to better execution since threads are not halted waiting for the surrounding neighbor cells to be computed when the amount of particles \RMchange{between} them is different.

@@

In order to attain distributed-memory parallelism and software flexibility while attaining good performance, P4IRS provides a customizable and robust communication module that facilitates integrating new domain partitioning schemes into its generated simulations.
This is done by combining its runtime functions and code generation techniques, and allows the usage of the waLBerla domain partitioning scheme based on a block-forest data structure.

@@

These methods make use of a domain partitioner class that can be implemented in the P4IRS runtime \RMchange{library} to determine which particles must be sent.
The method is implemented in C++ and fill\RMchange{s} domain buffers that are used for evaluating if a particle is outside the domain or within the border using its position.
This means that to include new domain partitioning schemes in P4IRS, it is only necessary to define a C++ runtime class, which can make use of external libraries and therefore can use the waLBerla block-forest domain partitioning.
Note that the reason for using buffers instead of evaluating the particle positions directly in the runtime code is performance, since particle packing and unpacking kernels should be executed in the target processor to achieve parallelism and to avoid copying data back and forth from the accelerator and host processors.

@@

Experiments for \ac{MD} use the standard Copper FCC lattice (see \autoref{fig:copper_fcc_lattice}) test-case from miniMD/MD-Bench, and we use our results from MD-Bench as a performance reference since it is a well optimized code for \ac{MD}.
The \ac{DEM} test case is a ``Settling Spheres'' benchmark (see \autoref{fig:settling_spheres}), on which several spheres fall into a ``ground'' (half-space), thus generating a ``bed'' of spherical particles.
Due to its multiple shapes and dynamics that consists of many interactions \RMchange{between} spheres and half-spaces, it should provide a good reference for DEM simulations.
We compare our single-core results from P4IRS with MESA-PD using 2 MPI ranks since it is not possible to run this setup in MESA-PD using a single rank.
GPU results are not available for MESA-PD as it is not capable of generating GPU code, but we show GPU performance results for P4IRS comparing different code generation strategies.

We also show weak-scaling results to demonstrate P4IRS scaling capabilities.
For this we use regular domain partitioning since these experiments do not require sophisticated load-balancing algorithms (Copper lattice is an homogeneous test-case in all dimensions, and Settling Spheres can obtain acceptable weak-scaling by regularly partitioning the domain in X and Y dimensions).
We use double floating-point precision and display results when enabling and disabling Newton's 3rd Law optimization (half neighbor-lists).
For \ac{MD} simulations, we show results with the Verlet Lists optimization strategy since it is widely employed in the field, while for \ac{DEM} we display results with the Linked Cells data-structure because particles represent rigid bodies that need to be in contact for their contributions to be computed.
Since the amount of contacts is typically way less than atomic interactions, there are no significant advantages of using the Verlet Lists data structure.

@@

For \RMchange{the} \ac{MD} case\RMchange{s}, we use the Verlet Lists algorithm as it is the standard one for such simulations.
The simulations runs through 200 time-steps, and the neighbor-lists are rebuilt at every 20 time-steps of the simulation (and at the start).
We use the simple \ac{LJ} potential (as shown in \autoref{lst:pairslj}) to calculate the force contributions \RMchange{between} atoms.

@@

\autoref{fig:md_weak_scaling_fritz} shows the results for P4IRS weak-scaling in the Fritz supercomputer.
For each node, a system of $128^3$ unit cells is included, with roughly 8.388.608 atoms per node (or 116.508 atoms per CPU core).
Since the \emph{multinode} partition in Fritz has the maximum of 64 nodes (which results in 4608 CPU cores, no hyper-threading), this is the maximum amount we scale up our problem to.
It is possible to see that we achieve perfect weak-scaling in all the settings we use, and the \ac{HN} version is significantly faster than the \ac{FN} version.
Notice that atomic operations are not required because we only make use of parallelism through spatial decomposition and message communication with MPI, thus we use distributed-memory parallelism strategy even within the same node and there is no risk of race conditions.
Based on our results, it is clear that P4IRS is able to generate scalable code on modern CPU clusters, but problems that are not too homogeneous may need proper mechanisms for balancing the workload \RMchange{between} different MPI ranks.
This is possible by using different domain partitioning methods in P4IRS such as the waLBerla domain partitioning with block-forests, but for this example we can achieve perfect scaling even with regular domain partitioning.

@@

\autoref{fig:md_weak_scaling_juwels} shows the weak-scaling runtimes for P4IRS in the JUWELS Booster cluster.
Each node simulates a system of $100^3$ unit cells, with roughly 500.000 atoms per GPU.
We scale up to 128 nodes, which is equivalent of 512 A100 GPUs.
Note that for the first node, the performance is slightly worse than the next ones, which can be explained as we only have 4 GPUs in one node, and domain partitioning is only done in two dimensions.
This can lead to worse performance because the surface area for communication is likely higher for the dimension where the partitioning is not done, which culminates in more atoms to be communicated and therefore more time spent packing and transferring data.
Apart from this, the scaling is also perfect for all the settings, and it is possible to observe that the \ac{FN} version is slightly faster than \ac{HN}.
As we already discussed, the usage of \ac{HN} is not necessarily beneficial in GPUs due to atomics, and this can be worse especially for kernels which do not exhibit high arithmetic intensity.
The results demonstrate that P4IRS is in fact capable of generating scalable code for modern GPU clusters.

@@

\autoref{fig:dem_single_core_runtimes} displays the runtimes for the \ac{DEM} kernels in single-core (P4IRS) and two cores/ranks (P4IRS/MESA-PD), also with cases building the neighbor-lists per cell (NPC) in P4IRS to evaluate its impact on CPU performance.
It is noticeable that P4IRS outperforms MESA-PD even with a single rank in the overall runtime, and what draws the attention is the amount of time MESA-PD spends executing other procedures instead of the force calculation and building the neighbor-lists
This can be attributed to the overhead that MESA-PD has to manage its domain partitioning with waLBerla (which cannot be switched) and calling its communication routines for next-neighbor communication.
A more in-depth analysis is required to actually understand why it takes a significant fraction of the time for MESA-PD to communicate data, but this is out of scope of this paper.
